---
title: "R Tutorial for GSRs"
author: "Randy Clopton"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview of RStudio
This document was created using RStudio, an IDE designed to make R easier to use
and to add in additional functionality. RStudio is the place you should go to 
get started on your R journey. To use RStudio, you do need to download R
separately, so make sure you've downloaded both.

You can download R here: https://www.r-project.org/

You can download RStudio here: https://posit.co/products/open-source/rstudio/

In addition, to use some of the key functionality that RStudio has built in, it 
may help to download a TeX editor (this just lets you use some formatting
features of RStudio better, and is totally optional), You can download the
preferred version for RStudio here: https://miktex.org/download

## RStudio Layout
RStudio has four main panes.
* Source pane: where you see your code
* Console pane: where you see your code run
* Environment pane: shows you what objects you've created, plus the code you've
run recently
* Output pane: shows you any files, plots, or other things you've created. Also
lets you find new packages, use the help feature, and enter presentation mode.

## R Script and Markdown Files
All of your work will be done in one of these two file types. R Scripts are just
documents where you put all of your code. R Markdown documents are a bit fuller
featured and let you do things like text formatting, organizing your code into
discrete chunks, and printing your output directly in the source pane. The main
advantage of markdown files for students is to create well formatted documents 
that are ready to use as your finalized problem set.

## RStudio Projects
RStudio also has a projects feature that lets you tie together a number of 
different files that are intended to be used in concert, including code, data 
files, and anything you output as part of your process. Creating a project also
tells R to load a directory that you created to contain all your files, which 
makes it easier to call up different files.

## Loading Relevant Packages
Before you start anything up in R, you will need to load in any packages you
want to use to conduct your analysis. We will discuss useful packages later, 
but for now, here's what the process of loading in packages looks like

```{r Load Packages}
#tidyverse is probably the most useful package and at least one of its elements
#will probably be loaded in every project you do
library(tidyverse)
#the plm package is a go-to for panel data. Once you get to fixed effects, this
#package will let you easily run models with defined fixed effects.
library(plm)
```


# Data Import and Manipulation (Ripper)

## Data Formats and Use Cases

### Data Types

### Data Structures

## Importing Data Files
```{r Import Data}

```


## Data Shaping and the Pipe Function

### Creating a dummy variable (RC Note: this is a specific thing that needs to be in this section as it's what is going to get us to the analysis part of the document)

# Conducting Statistical Analyses Using R (Clopton)

## Generating Summary Statistics
```{r}

```

## T Tests and Correlations
```{r}

```

## Regressions
```{r}

```

## F Tests
```{r}

```

## Interaction Terms and Difference-in-Difference
```{r}

```

## Visualizations
```{r}

```

## Additional Useful Packages
